From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2017-11-01 paper.Rmd
hard-wrap lines

---
title: "From noise to knowledge: how randomness generates novel phenomena and information"
author:
  - name: "Carl Boettiger"
    affiliation: a
    email: "cboettig@berkeley.edu"

@@

@@

Highlight *key questions* & *key advances*
Moving beyond "everything is important" [@Bjornstad2001; @Kingsland1985]

(@Bjornstad2001 future: more MCMC methods. More mechanistic models.  dealing with unobserved dynamics)

Shift from noise as nuiance to noise as opportunity
Stochastic phenomena outside of theory -- early warning signals

Bialek's rules
1. Noise isn't negligible: photon counting, chemotaxis, 
2. No fine tuning: developmental patterning? ion channels/ stochastic switches? 
3. Efficient representation: entropy, information flow

@@

<!--
Motivate the presentation of the van Kampen result rather than a generic review.
-->

@@

The importance of $\tau_c$, or the degree of auto-correlation, which emerges from our derivation above, has also been a subject of significant interest in the study of stochastic population dynamics [e.g.@Rougharden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].   Early work rarely often assumed environmental noise would be uncorrelated as a mathematical convenience.  As the derivation of Eq \eqref{env_noise} makes clear, for a continuous-time process, the relevant metric is the timescale of auto-correlation in the environmental noise process relative to the timescale (auto-correlation) of the population dynamics.  The Gaussian noise that emerges out of the system-size approximation is always auto-correlated to some degree -- if we measure our environment at consecutive times infinitesimally far apart, we expect nearly identical values.  At intervals much longer than the correlation time, consecutive measurements are effectively independent: auto-correlated noise is all about _relative_ timescales.

@@

<!-- 
Noise is "tacked on" in these models... contrasts to previous section
-->

@@

A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species. Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guaranteed to eventually go extinct under demographic noise.  While extinction is certainly a phenomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself. Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.  In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibrium of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].

@@

**Bet hedging**
**Inflation**

@@

$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$
where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$. Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.  Equilibria exist whenever growth balances consumption.  For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.  When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.  Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.  Stochastic fluctuations drive spontaneous shifts between these alternative stable states.

>>>>

---
title: "From noise to knowledge: how randomness generates novel phenomena and reveals information"
author:
  - name: "Carl Boettiger"
    affiliation: a
    email: "cboettig@berkeley.edu"

@@

<!--
Directions:

@@

@@

@@

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work rarely often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the timescale (auto-correlation)
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent: auto-correlated noise is
all about _relative_ timescales.

@@

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise.  While extinction
is certainly a phenomenon that stands in stark contrast to the
long-term behavior of the corresponding deterministic equation,
this result is not particularly useful: "eventually" could be longer
than the existence of the planet. Rather, it is a reminder that we
must be careful about how we describe the 'long-term' behavior of
a model.  In place of the mathematical notion of a stationary
distribution to describe the long-term behavior of a stochastic
model, analysis of demographic stochasticity focuses on the concept
of quasi-stationary distributions, which in many cases correspond
more closely to the equilibrium of their deterministic counterparts
[@Ovaskainen2010; @Schreiber2017].

@@

**Bet hedging**

@@

$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$

where $\xi_t$ is normally distributed environmental variation with
mean zero and standard deviation $\sigma$. Figure 3B shows separate
curves corresponding the parts of the model describing vegetation
growth and vegetation consumption, using the same parameters as
shown in the simulation panel, 3A.  Equilibria exist whenever growth
balances consumption.  For the parameters shown, the S-shaped curve
of the Type-III consumption pattern crosses the quadratic curve of
logistic growth at four separate locations.  When growth is larger
than consumption before the crossing, and thus smaller after, the
point is stable, otherwise it is unstable.  Consequently, we can
see two stable points, in the neighborhood of $X_t \approx 1.2$ and
$X_t \approx 0.5$, closely matching the average fluctuations seen
in the simulation in 3A.  Stochastic fluctuations drive spontaneous
shifts between these alternative stable states.
